\textbf{Urban mobility and socioeconomic deprivation in Latin America after COVID-19}

### 2024-02-06 manuscript/manuscript.tex
add refs to intro

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study, we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities are present
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. These disparities generally diminish over
time, although they persist in all the countries included in the
analysis. Our research highlights the importance of timely mobility data
with high spatiotemporal resolution to understand the long-term effects
of the pandemic and to inform equitable policy responses that address
societal challenges in urban areas.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Spatial human mobility is key to creating sustainable, livable and
inclusive cities. At the societal level, spatial mobility enables the
transfer of knowledge, skills and labour to places they are needed.
Spatial mobility also shapes service and transport demand across urban
spaces, and enables the monitoring and control of transmissible
diseases. At the individual level, mobility enables people to access and
achieve opportunities and aspirations in space. Understanding spatial
human mobility is thus important to supporting appropriate policy
responses to address societal challenges relating to carbon emissions,
urban planning, service delivery, public health, disaster management and
transport.

The COVID-19 pandemic resulted in a notable decrease in mobility,
particularly in cities. This decrease in the overall levels of mobility
was prompted by nonpharmaceutical interventions to contain the spread of
COVID‐19, coupled with general fears of visiting crowded public
spaces.Especially during lockdowns, mobility recorded reductions in the
frequency, distance and time of trips in cities across the globe. Higher
engagement with remote working, online schooling and shopping activity
reduced the need to travel for work, education, shopping and leisure,
hence giving rise to more geographically localised mobility patterns.

However, reductions in mobility levels were highly unequal reflecting
existing socioeconomic inequalities. In most countries, affluent
individuals tended to record the greatest drops in mobility levels as
they are predominantly employed in knowledge-intensive jobs which can be
done fully or partly remotely. During the COVID-19 pandemic, the
adoption of remote work reduced the need of commuting for
knowledge-intensive, non-public facing jobs. At the same time,
individuals from less privileged socioeconomic backgrounds displayed
less pronounced declines mirroring the nature of their work requiring
public-facing, face-to-face interaction, and thus a requirement for
daily work commutes.

Thus, while a growing body of empirical evidence has contributed to
advancing our understanding of the impacts of the COVID-19 pandemic on
spatial mobility within cities, existing research has focused on more
developed countries and the immediate effects of the pandemic during
2020. Less is known about the longer term patterns of resilience in
urban mobility in less developed countries. Assessing the extent to
which the level of mobility has returned back to the pre-pandemic
baseline level across socioeconomic groups is important to understand
the potentially unequal impacts of hybrid working in the society.

A key barrier to monitor changes in geographic mobility patterns in less
developed countries during and post the COVID-19 pandemic has been the
lack of suitable data. Traditionally census and survey data have been
employed to analyse human mobility patterns in these countries. Yet,
these data streams are not frequently updated and suffer from slow
releases, with census data for example being collected over intervals of
ten years in most countries. Traditional data streams thus lack the
temporal granularity to analyse population movements over short-time
periods and to offer an up-to-date representation of the urban mobility
system. Data resulting from social interactions on digital platforms
have emerged as an unique source of information to deliver this
representation and capture human population movement in less developed
countries at scale. Particularly location data from mobile phone
applications have become a prominent source to sense patterns of human
mobility at higher geographical and temporal resolution in real time.

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL), developed by the European Comission's Joint Research Centre
(Schiavina M. 2019) to define urban areas; and the Global Gridded
Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data
and Applications Centre (Columbia University 2022) from sociodemographic
and satellite data inputs. Building on existing evidence (e.g. (Rowe et
al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
has recovered returning to the pre-pandemic baseline level of movement
as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive businesses adopt hybrid
working.

\begin{itemize}
\tightlist
\item
  Context and importance of human mobility: public health , climate
  change\\
\item
  The pandemic resulted in major changes in human mobility - unequal
  impacts across socioeconomic groups\\
\item
  Gap 1: Little work on understanding the long-term changes in mobility
  - recovery\\
\item
  Gap 2: little work on less developed countries - where inequalities
  are more pronounced\\
\item
  Aim - focus on Latin American countries as case study.
\item
  Aim - focus on urban areas, as the future of humanity is the future of
  cities and Latin Ameriaca is highly urbanised
\end{itemize}

@@

The evolution of the percentage change in the number of movements is
measured with respect to a baseline period prior to the pandemic as
described in Section~\ref{sec-methods}. For the purposes of the
analysis, we aggregate the raw movement data temporally into months and
spatially into administrative units at various levels according to GADM,
the Database of Global Administrative Areas (GADM 2024). The analysis
focuses on administrative units that are within the boundaries of
functional urban areas as specified by the Global Human Settlement Layer
(GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number
of movements are made using the Prophet forecasting procedure. Further
details are provided in Section~\ref{sec-methods}.

@@

We analyse the evolution of the percentage change in the number of
movements with respect to a baseline period prior to the pandemic.
Specifically, we focus the analysis on short-distance movements in urban
areas to represent local and routine mobility (Owen and Green 1992), so
only movements covering a distance of at most 70 km are considered. For
a movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

@@

We focused the analysis on short-distance movements in urban areas,
especifically those covering 70 km or less. These journeys are typically
considered to represent local and routine mobility (Owen and Green 1992)
. However, due to the characteristics of the Meta-Facebook movement
data, we are unable to distinguish the purpose of these short-distance
movements. Hence, some of our data could be capturing journeys that
involve a permanent change of place of residence. Our work therefore
motivates the need to answer questions regarding the validity of digital
footprint data for the analysis of human mobility. Further research
should focus on inferring more specific information about the nature of
the journeys, following similar approaches to those proposed by
Cabrera-Arnau et al. (2023), and quantifying the extent to which the
digital footprint data mirrors the true mobility patterns.

Conducting research on urban mobility using digital footprint data is
not straightforward, due to the challenges in accessing and working with
unstructured data sets which are often subject to biases and statistical
representation issues. These biases often arise from inequalities in
access and usage of digital technologies across demographic groups (Rowe
2023). Despite these challenges, the data and analysis that we used for
this work provide evidence for non-trivial patterns that are consistent
across four countries in Latin America and with other parts of the
world. Our findings highlight the dynamic interplay between
socioeconomic status and urban mobility, and shall be used to motivate
and inform the public debate regarding the deep societal consequences of
urban mobility disparities on the wider socioeconomic landscape of Latin
American countries.

In conclusion, we argue that this work goes beyond the analysis of
specific patterns by demonstrating the potential of digital footprint
data for policy-relevant research on human mobility at an unprecedented
level of spatiotemporal granularity. While we have seen a rise in
initiatives to improve data services and methodological frameworks to
facilitate the use of digital footprint data for social good, progress
is still limited, especially in some parts of the world including Latin
America. It is in the hands of governments and public organisations to
prioritise the maximisation the societal benefits that digital footprint
data has to offer. This includes engaging in activities such as building
strategic partnerships with commercial data-holders and academic
institutions to establish a unified framework for the use of digital
footprint data in policy and research. In particular, we call for the
creation of resources like those developed by the European Commission
Joint Research Centre (Commission et al. 2022) and the UN Statistics
Division (Division 2019), which identify sources of non-traditional data
and set methodological protocols for incorporating mobile phone data
into official mobility statistics. While current resources tend to have
a global reach, we advocate for more tailored local initiatives that
acknowledge disparities in regional data availability and adoption of
digital technologies.

@@

To capture population movements, we used anonymised aggregate mobile
phone location data from Meta users for Argentina, Colombia, Chile and
Mexico, covering a 24-month period from April 2020 to March 2022. We
used the dataset Facebook Movements created by Meta and accessed through
their Data for Good Initiative (\url{https://dataforgood.facebook.com}).
The data are built from Facebook app users who have the location
services setting turned on on their smartphone. Prior to releasing the
datasets, Meta ensures privacy and anonymity by removing personal
information and applying privacy-preserving techniques (Maas et al.
2019). Small-count dropping is one of these techniques. A data entry is
removed if the population or movement count for an area is lower than
10. The removal of small counts may mean that population counts in small
sparsely populated areas are not captured. A second technique consists
in adding a small undisclosed amount of random noise to ensure that it
is not possible to ascertain precise, true counts for sparsely populated
locations. Third, spatial smoothing using inverse distance-weighted
averaging is also applied is applied to produce a smooth population
count surface. The Facebook Movements dataset offers information on the
total number of Facebook users moving between and within spatial units
in the form of origin-destination matrices. The data is temporally
aggregated into three daily 8-hour time windows (i.e.~00:00-08:00,
08:00-16:00 and 16:00-00:00). The dataset includes a baseline capturing
the number of movements before COVID-19 based on a 45-day period ending
on March 10th 2020. The baseline is computed using an average for the
same time of the day and day of the week in the period preceding March
10th. For instance, the baseline for Monday 00:00-08:00 time window is
obtained by averaging over data collected on Mondays from 00:00 to 8:00
for the 45-day period. Details about the baseline can be found in (Maas
et al. 2019). The Bing Maps Tile System developed by Microsoft
(Microsoft) is used a spatial framework to organise the data. The Tile
System is a geospatial indexing system that partitions the world into
tile cells in a hierarchical way, comprising 23 different levels of
detail (Microsoft). At the lowest level of detail (Level 1), the world
is divided into four tiles with a coarse spatial resolution. At each
successive level, the resolution increases by a factor of two. The data
that we used are spatially aggregated into Bing tile levels 13. That is
about 4.9 x 4.9km at the Equator (Maas et al. 2019).

@@

For the countries included in the analysis, the movement data is
available until March 2022. In order to gain a better understanding of
the recovery trends after the pandemic, we generate a 12-month forecast
up to March 2023. This is done using Prophet (Taylor and Letham 2017), a
procedure developed by Meta's Core Data Science team to forecast time
series data based on an additive model. Prophet stands out from other
forecasting methods due to its ability to capture recurring seasonal
trends inherent in mobility data, such as fluctuations during holidays,
or changes in the general trend due to the impact of particular
circumstances, such as the COVID-19 pandemic. It is particularly
intuitive to use compared to other models due to its automated
trend-detection capabilities, making it accessible to users with no
expertise on the underlying model, while being robust to missing data
and outliers. For the analysis in this paper, yearly seasonality effects
are included as they yield the most realistic predictions.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-LaurenAlexander15}{}}%
Alexander, Lauren, Shan Jiang, Mikel Murga, and Marta C. González. 2015.
{``Origin--Destination Trips by Purpose and Time of Day Inferred from
Mobile Phone Data.''} \emph{Transportation Research Part C: Emerging
Technologies} 58: 240--50.
https://doi.org/\url{https://doi.org/10.1016/j.trc.2015.02.018}.

\leavevmode\vadjust pre{\hypertarget{ref-Cabrera-Arnau2023}{}}%
Cabrera-Arnau, Carmen, Chen Zhong, Michael Batty, Ricardo Silva, and
Soong Moon Kang. 2023. {``Inferring Urban Polycentricity from the
Variability in Human Mobility Patterns.''} \emph{Scientific Reports} 13
(1): 5751. \url{https://doi.org/10.1038/s41598-023-33003-7}.

\leavevmode\vadjust pre{\hypertarget{ref-RDI2022}{}}%
Columbia University, Center for International Earth Science Information
Network. CIESIN. 2022. {``Global Gridded Relative Deprivation Index
(GRDI), Version 1.''} Palisades, New York: NASA Socioeconomic Data;
Applications Center (SEDAC). \url{https://doi.org/10.7927/3xxe-ap97}.

\leavevmode\vadjust pre{\hypertarget{ref-ECJRC22}{}}%
Commission, European, Joint Research Centre, C Bosco, S
Grubanov-Boskovic, S Iacus, U Minora, F Sermi, and S Spyratos. 2022.
\emph{Data Innovation in Demography, Migration and Human Mobility}.
Publications Office of the European Union.
\url{https://doi.org/doi/10.2760/027157}.

\leavevmode\vadjust pre{\hypertarget{ref-UNSD}{}}%
Division, United Nations Statistical. 2019. {``Handbook on the Use of
Mobile Phone Data for Official Statistics.''}

\leavevmode\vadjust pre{\hypertarget{ref-GADM24}{}}%
GADM. 2024. {``GADM Maps and Data.''}

\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\leavevmode\vadjust pre{\hypertarget{ref-MengChuishi17}{}}%
Meng, Chuishi, Yu Cui, Qing He, Lu Su, and Jing Gao. 2017. {``Travel
Purpose Inference with GPS Trajectories, POIs, and Geo-Tagged Social
Media Data.''} In \emph{2017 IEEE International Conference on Big Data
(Big Data)}, 1319--24.
\url{https://doi.org/10.1109/BigData.2017.8258062}.

\leavevmode\vadjust pre{\hypertarget{ref-NiluferSariAslam21}{}}%
Nilufer Sari Aslam, Tao Cheng, Mohamed R. Ibrahim, and Yang Zhang. 2021.
{``ActivityNET: Neural Networks to Predict Public Transport Trip
Purposes from Individual Smart Card Data and POIs.''} \emph{Geo-Spatial
Information Science} 24 (4): 711--21.
\url{https://doi.org/10.1080/10095020.2021.1985943}.

\leavevmode\vadjust pre{\hypertarget{ref-owen1992migration}{}}%
Owen, D., and A. Green. 1992. {``Migration Patterns and Trends.''} In
\emph{Migration Processes and Patterns: Research Progress and
Prospects}, edited by T. Champion and T. Fielding. Belhaven Press.

\leavevmode\vadjust pre{\hypertarget{ref-RoweBigData23}{}}%
Rowe, Francisco. 2023. {``9.: Big Data.''} In \emph{Concise Encyclopedia
of Human Geography}, 42--47. Cheltenham, UK: Edward Elgar Publishing.
\url{https://doi.org/10.4337/9781800883499.ch09}.

\leavevmode\vadjust pre{\hypertarget{ref-roweCalafiore2022}{}}%
Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
Understanding Human Mobility in Britain During the COVID-19 Pandemic
Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.

\leavevmode\vadjust pre{\hypertarget{ref-GHSL19}{}}%
Schiavina M., Maffenini L., Moreno-Monroy A. 2019. {``GHS-FUA R2019A -
GHS Functional Urban Areas, Derived from GHS-UCDB R2019A, (2015),
R2019A.''} European Commission, Joint Research Centre (JRC).

\leavevmode\vadjust pre{\hypertarget{ref-prophet17}{}}%
Taylor, Sean J, and Benjamin Letham. 2017. {``Forecasting at Scale.''}
\emph{PeerJ Preprints} 5 (September): e3190v2.
\url{https://doi.org/10.7287/peerj.preprints.3190v2}.

>>>>

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study, we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities are present
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. These disparities generally diminish over
time, although they persist in all the countries included in the
analysis . Our research highlights the importance of timely mobility
data with high spatiotemporal resolution to understand the long-term
effects of the pandemic and inform equitable policy responses that
address societal challenges in urban areas.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Spatial human mobility is key to creating sustainable, livable and
inclusive cities. At the societal level, spatial mobility enables the
transfer of knowledge, skills and labour to places they are
needed\textsuperscript{1}. Spatial mobility also shapes service and
transport demand across urban spaces\textsuperscript{2}, and enables the
monitoring and control of transmissible diseases\textsuperscript{3}. At
the individual level, mobility enables people to access and achieve
opportunities and aspirations in space\textsuperscript{4}. Understanding
spatial human mobility is thus important to supporting appropriate
policy responses to address societal challenges relating to carbon
emissions, urban planning, service delivery, public health, disaster
management and transport\textsuperscript{5,6}.

The COVID-19 pandemic resulted in a notable decrease in mobility,
particularly in cities\textsuperscript{7}. Coupled with fears of
contagion in crowded public spaces, nonpharmaceutical interventions to
contain the spread of COVID‐19 prompted this decrease in the overall
levels of mobility\textsuperscript{7,8}. Especially during lockdowns,
mobility recorded reductions in the frequency, distance and time of
trips in cities across the globe\textsuperscript{9--12}. Higher
engagement with remote working, online schooling and shopping activity
reduced the need to travel for work, education, shopping and leisure,
hence giving rise to more geographically localised mobility
patterns\textsuperscript{13}.

However, reductions in mobility levels were highly unequal reflecting
existing socioeconomic inequalities in our
societies\textsuperscript{14}. In most countries, affluent individuals
tended to record the greatest drops in mobility levels as they are
predominantly employed in knowledge-intensive jobs which can be done
fully or partly remotely\textsuperscript{10,15--18}. During the COVID-19
pandemic, the adoption of remote work reduced the need of commuting for
knowledge-intensive, non-public facing jobs\textsuperscript{19}. At the
same time, individuals from less privileged socioeconomic backgrounds
displayed less pronounced declines mirroring the nature of their work
requiring public-facing, face-to-face interaction, and thus a
requirement for daily work commutes\textsuperscript{17,18}.

Thus, while a growing body of empirical evidence has contributed to
advancing our understanding of the impacts of the COVID-19 pandemic on
spatial mobility within cities, existing research has focused on more
developed countries and the immediate effects of the pandemic during
2020. Less is known about the longer term patterns of resilience in
urban mobility in less developed countries extending beyond this
period\textsuperscript{20}. Urban spaces have changed considerably since
then, from going through waves of high COVID-19 fatality, infections,
school and business closures to the removal of all COVID-19 restrictions
as the UN World Health Organization (WHO) declared an end to the
pandemic as a public health emergency; yet, different configurations of
hybrid working have remained in the norm across many sector of the
economy\textsuperscript{21,22}. Thus, assessing the extent to which the
level of mobility has returned back to the pre-pandemic baseline level
across socioeconomic groups is important to understand the potentially
unequal long-term impacts of hybrid working.

A key barrier to monitor changes in geographic mobility patterns in less
developed countries during and post the COVID-19 pandemic has been the
lack of suitable data\textsuperscript{20}. Traditionally census and
survey data have been employed to analyse human mobility patterns in
these countries\textsuperscript{23}. Yet, these data streams are not
frequently updated and suffer from slow releases, with census data for
example being collected over intervals of ten years in most
countries\textsuperscript{24}. Traditional data streams thus lack the
temporal granularity to analyse population movements over short-time
periods and to offer an up-to-date representation of the urban mobility
system\textsuperscript{25}. Data resulting from social interactions on
digital platforms have emerged as an unique source of information to
deliver this representation and capture human population movement in
less developed countries at scale\textsuperscript{25}. Particularly
location data from mobile phone applications have become a prominent
source to sense patterns of human mobility at higher geographical and
temporal resolution in real time\textsuperscript{26}.

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm\textsuperscript{27}
to predict origin-destination flows from June 2022 to March 2023. We use
Functional Urban Areas (FUAs) boundaries from the Global Human
Settlement Layer (GHSL), developed by the European Comission's Joint
Research Centre\textsuperscript{28} to define urban areas; and the
Global Gridded Relative Deprivation Index (GRDI) developed by NASA's
Socioeconomic Data and Applications Centre\textsuperscript{29} from
sociodemographic and satellite data inputs. Building on existing
evidence\textsuperscript{e.g. 30,31}, we hypothesised that (1) urban
mobility has recovered returning to the pre-pandemic baseline level of
movement as nonpharmaceuthical restrictions were lifted; and, that (2)
socioeconomic differences in urban mobility have endured the pandemic
reflecting deep societal inequalities as knowledge-intensive businesses
adopt hybrid working. \\

\begin{itemize}
\tightlist
\item
  Aim - focus on Latin American countries as case study.
\item
  Aim - focus on urban areas, as the future of humanity is the future of
  cities and Latin Ameriaca is highly urbanised
\end{itemize}

@@

The evolution of the percentage change in the number of movements is
measured with respect to a baseline period prior to the pandemic as
described in Section~\ref{sec-methods}. For the purposes of the
analysis, we aggregate the raw movement data temporally into months and
spatially into administrative units at various levels according to GADM,
the Database of Global Administrative Areas\textsuperscript{32}. The
analysis focuses on administrative units that are within the boundaries
of functional urban areas as specified by the Global Human Settlement
Layer (GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number
of movements are made using the Prophet forecasting procedure. Further
details are provided in Section~\ref{sec-methods}.

@@

We analyse the evolution of the percentage change in the number of
movements with respect to a baseline period prior to the pandemic.
Specifically, we focus the analysis on short-distance movements in urban
areas to represent local and routine mobility\textsuperscript{33}, so
only movements covering a distance of at most 70 km are considered. For
a movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

@@

We focused the analysis on short-distance movements in urban areas,
especifically those covering 70 km or less. These journeys are typically
considered to represent local and routine mobility\textsuperscript{33} .
However, due to the characteristics of the Meta-Facebook movement data,
we are unable to distinguish the purpose of these short-distance
movements. Hence, some of our data could be capturing journeys that
involve a permanent change of place of residence. Our work therefore
motivates the need to answer questions regarding the validity of digital
footprint data for the analysis of human mobility. Further research
should focus on inferring more specific information about the nature of
the journeys, following similar approaches to those proposed
by\textsuperscript{37}, and quantifying the extent to which the digital
footprint data mirrors the true mobility patterns.

Conducting research on urban mobility using digital footprint data is
not straightforward, due to the challenges in accessing and working with
unstructured data sets which are often subject to biases and statistical
representation issues. These biases often arise from inequalities in
access and usage of digital technologies across demographic
groups\textsuperscript{38}. Despite these challenges, the data and
analysis that we used for this work provide evidence for non-trivial
patterns that are consistent across four countries in Latin America and
with other parts of the world. Our findings highlight the dynamic
interplay between socioeconomic status and urban mobility, and shall be
used to motivate and inform the public debate regarding the deep
societal consequences of urban mobility disparities on the wider
socioeconomic landscape of Latin American countries.

In conclusion, we argue that this work goes beyond the analysis of
specific patterns by demonstrating the potential of digital footprint
data for policy-relevant research on human mobility at an unprecedented
level of spatiotemporal granularity. While we have seen a rise in
initiatives to improve data services and methodological frameworks to
facilitate the use of digital footprint data for social good, progress
is still limited, especially in some parts of the world including Latin
America. It is in the hands of governments and public organisations to
prioritise the maximisation the societal benefits that digital footprint
data has to offer. This includes engaging in activities such as building
strategic partnerships with commercial data-holders and academic
institutions to establish a unified framework for the use of digital
footprint data in policy and research. In particular, we call for the
creation of resources like those developed by the European Commission
Joint Research Centre\textsuperscript{39} and the UN Statistics
Division\textsuperscript{40}, which identify sources of non-traditional
data and set methodological protocols for incorporating mobile phone
data into official mobility statistics. While current resources tend to
have a global reach, we advocate for more tailored local initiatives
that acknowledge disparities in regional data availability and adoption
of digital technologies.

@@

To capture population movements, we used anonymised aggregate mobile
phone location data from Meta users for Argentina, Colombia, Chile and
Mexico, covering a 24-month period from April 2020 to March 2022. We
used the dataset Facebook Movements created by Meta and accessed through
their Data for Good Initiative (\url{https://dataforgood.facebook.com}).
The data are built from Facebook app users who have the location
services setting turned on on their smartphone. Prior to releasing the
datasets, Meta ensures privacy and anonymity by removing personal
information and applying privacy-preserving
techniques\textsuperscript{41}. Small-count dropping is one of these
techniques. A data entry is removed if the population or movement count
for an area is lower than 10. The removal of small counts may mean that
population counts in small sparsely populated areas are not captured. A
second technique consists in adding a small undisclosed amount of random
noise to ensure that it is not possible to ascertain precise, true
counts for sparsely populated locations. Third, spatial smoothing using
inverse distance-weighted averaging is also applied is applied to
produce a smooth population count surface. The Facebook Movements
dataset offers information on the total number of Facebook users moving
between and within spatial units in the form of origin-destination
matrices. The data is temporally aggregated into three daily 8-hour time
windows (i.e.~00:00-08:00, 08:00-16:00 and 16:00-00:00). The dataset
includes a baseline capturing the number of movements before COVID-19
based on a 45-day period ending on March 10th 2020. The baseline is
computed using an average for the same time of the day and day of the
week in the period preceding March 10th. For instance, the baseline for
Monday 00:00-08:00 time window is obtained by averaging over data
collected on Mondays from 00:00 to 8:00 for the 45-day period. Details
about the baseline can be found in\textsuperscript{41}. The Bing Maps
Tile System developed by Microsoft (Microsoft) is used a spatial
framework to organise the data. The Tile System is a geospatial indexing
system that partitions the world into tile cells in a hierarchical way,
comprising 23 different levels of detail (Microsoft). At the lowest
level of detail (Level 1), the world is divided into four tiles with a
coarse spatial resolution. At each successive level, the resolution
increases by a factor of two. The data that we used are spatially
aggregated into Bing tile levels 13. That is about 4.9 x 4.9km at the
Equator\textsuperscript{41}.

@@

For the countries included in the analysis, the movement data is
available until March 2022. In order to gain a better understanding of
the recovery trends after the pandemic, we generate a 12-month forecast
up to March 2023. This is done using Prophet\textsuperscript{27}, a
procedure developed by Meta's Core Data Science team to forecast time
series data based on an additive model. Prophet stands out from other
forecasting methods due to its ability to capture recurring seasonal
trends inherent in mobility data, such as fluctuations during holidays,
or changes in the general trend due to the impact of particular
circumstances, such as the COVID-19 pandemic. It is particularly
intuitive to use compared to other models due to its automated
trend-detection capabilities, making it accessible to users with no
expertise on the underlying model, while being robust to missing data
and outliers. For the analysis in this paper, yearly seasonality effects
are included as they yield the most realistic predictions.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-ackers2005}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Ackers, L.
\href{https://doi.org/10.1111/j.1468-2435.2005.00343.x}{Moving People
and Knowledge: Scientific Mobility in the European Union1}.
\emph{International Migration} \textbf{43}, 99--131 (2005).}

\leavevmode\vadjust pre{\hypertarget{ref-chen2016}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Chen, C., Ma, J., Susilo, Y., Liu, Y. \& Wang, M.
\href{https://doi.org/10.1016/j.trc.2016.04.005}{The promises of big
data and small data for travel behavior (aka human mobility) analysis}.
\emph{Transportation Research Part C: Emerging Technologies}
\textbf{68}, 285--299 (2016).}

\leavevmode\vadjust pre{\hypertarget{ref-belik2011}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Belik, V., Geisel, T. \& Brockmann, D.
\href{https://doi.org/10.1103/physrevx.1.011001}{Natural Human Mobility
Patterns and Spatial Spread of Infectious Diseases}. \emph{Physical
Review X} \textbf{1}, (2011).}

\leavevmode\vadjust pre{\hypertarget{ref-klugman2009human}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Klugman, J. Human development report 2009. Overcoming
barriers: Human mobility and development. \emph{Overcoming Barriers:
Human Mobility and Development (October 5, 2009). UNDP-HDRO Human
Development Reports} (2009).}

\leavevmode\vadjust pre{\hypertarget{ref-barbosa2018}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Barbosa, H. \emph{et al.}
\href{https://doi.org/10.1016/j.physrep.2018.01.001}{Human mobility:
Models and applications}. \emph{Physics Reports} \textbf{734}, 1--74
(2018).}

\leavevmode\vadjust pre{\hypertarget{ref-chinazzi2020}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Chinazzi, M. \emph{et al.}
\href{https://doi.org/10.1126/science.aba9757}{The effect of travel
restrictions on the spread of the 2019 novel coronavirus (COVID-19)
outbreak}. \emph{Science} \textbf{368}, 395--400 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Nouvellet, P. \emph{et al.}
\href{https://doi.org/10.1038/s41467-021-21358-2}{Reduction in mobility
and COVID-19 transmission}. \emph{Nature Communications} \textbf{12},
(2021).}

\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Rowe, F., González-Leonardo, M. \& Champion, T. Virtual
special issue: Internal migration in times of COVID{-}19.
\emph{Population, Space and Place} (2023)
doi:\href{https://doi.org/10.1002/psp.2652}{10.1002/psp.2652}.}

\leavevmode\vadjust pre{\hypertarget{ref-abdullah2020}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Abdullah, M., Dias, C., Muley, D. \& Shahin, Md.
\href{https://doi.org/10.1016/j.trip.2020.100255}{Exploring the impacts
of COVID-19 on travel behavior and mode preferences}.
\emph{Transportation Research Interdisciplinary Perspectives}
\textbf{8}, 100255 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-bonaccorsi2020}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Bonaccorsi, G. \emph{et al.}
\href{https://doi.org/10.1073/pnas.2007658117}{Economic and social
consequences of human mobility restrictions under COVID-19}.
\emph{Proceedings of the National Academy of Sciences} \textbf{117},
15530--15535 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-abu-rayash2020}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Abu-Rayash, A. \& Dincer, I.
\href{https://doi.org/10.1016/j.erss.2020.101693}{Analysis of mobility
trends during the COVID-19 coronavirus pandemic: Exploring the impacts
on global aviation and travel in selected cities}. \emph{Energy Research
\& Social Science} \textbf{68}, 101693 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-lee2023}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Lee, S., Ko, E., Jang, K. \& Kim, S.
\href{https://doi.org/10.1016/j.cities.2023.104223}{Understanding
individual-level travel behavior changes due to COVID-19: Trip
frequency, trip regularity, and trip distance}. \emph{Cities}
\textbf{135}, 104223 (2023).}

\leavevmode\vadjust pre{\hypertarget{ref-borkowski2021}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Borkowski, P., Jażdżewska-Gutta, M. \& Szmelter-Jarosz,
A. \href{https://doi.org/10.1016/j.jtrangeo.2020.102906}{Lockdowned:
Everyday mobility changes in response to COVID-19}. \emph{Journal of
Transport Geography} \textbf{90}, 102906 (2021).}
